---
title: "Cyclistic Bike Share Case Study"
author: "Kengne Romaric"
date: "2025-12-11"
output: html_document
---

### Google Data Analytics Capstone project


## I.Context

Cyclistic, is a bike-share company in Chicago.

The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. 

Therefore, our team wants to understand how casual riders and annual members use Cyclistic bikes differently.

As a junior data analyst working on the marketing analyst team at Cyclistic, I was tasked to analyze the Cyclistic historical bike trip data to identify trends.

From these insights, our team will design a new marketing strategy to convert casual riders into annual members.

NB:

*Customers who purchase single-ride or full-day passes are referred to as casual riders. 

*Customers who purchase annual memberships are Cyclistic members.

## II.Task

How differently do casual riders and annual members use cyclistic bikes?

**Stakeholder:**

*Director of marketing/Manager

*The marketing analytics team

*The executive team

## III.Data source

**Source:** https://divvy-tripdata.s3.amazonaws.com/index.html 

Data provided by Motivate International Inc.

Our data set is constituted of rides from the Past 12 months on 08/12/2025.

The data is constituted of rides from 11/2024 to 10/2025.

## IV.Preparation of the dataset

All 12 months datasets were downloaded locally from the source cited above and analyzed using RStudio using R 4.5.2

We began our analysis by loading the Tidyverse package
```{r loading the tidyverse package}
library(tidyverse)
```

We set our local file as working directory
```{r setting working directory}
#Setting our local file containing the datasets as local directory
setwd("E:/DOCUMENTS ROMARIC/Data analytics/Google data analytics certification/Resources/Course 8/Case study 1/divvy trip data")
```

We imported all data sets into RStudio
```{r importing datasets}
#Importing all 12 datasets into RStudio
x202411 <- read.csv("202411-divvy-tripdata.csv")
x202412 <- read.csv("202412-divvy-tripdata.csv")
x202501 <- read.csv("202501-divvy-tripdata.csv")
x202502 <- read.csv("202502-divvy-tripdata.csv")
x202503 <- read.csv("202503-divvy-tripdata.csv")
x202504 <- read.csv("202504-divvy-tripdata.csv")
x202505 <- read.csv("202505-divvy-tripdata.csv")
x202506 <- read.csv("202506-divvy-tripdata.csv")
x202507 <- read.csv("202507-divvy-tripdata.csv")
x202508 <- read.csv("202508-divvy-tripdata.csv")
x202509 <- read.csv("202509-divvy-tripdata.csv")
x202510 <- read.csv("202510-divvy-tripdata.csv")
```

 We merged all the datasets into one.
 
 But before,
 
 We checked all columns within the datasets
```{r checking column names}
#Comparing column names
colnames(x202411)
colnames(x202412)
colnames(x202501)
colnames(x202502)
colnames(x202503)
colnames(x202504)
colnames(x202505)
colnames(x202506)
colnames(x202507)
colnames(x202508)
colnames(x202509)
colnames(x202510)
```
 
We then checked each column data type
```{r checking column data types}
str(x202411)
str(x202412)
str(x202501)
str(x202502)
str(x202503)
str(x202504)
str(x202505)
str(x202506)
str(x202507)
str(x202508)
str(x202509)
str(x202510)
```

We aggregated all the datasets into one
```{r aggregating all datasets}
#Aggregating all datasets into one
divvy_bike_trips <- bind_rows(x202411,x202412,x202501,x202502,x202503,x202504,x202505,x202506,x202507,x202508,x202509,x202510)

```
```{r View aggregated dataset}
head(divvy_bike_trips)
```

We filtered out all unnecessary columns
```{r filtering columns}
divvy_bike_trips <- divvy_bike_trips %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
```
```{r view filtered dataset}
head(divvy_bike_trips)
```

We inspected our new dataframe
```{r inspecting our dataframe}
#List of column names
colnames(divvy_bike_trips)  
#How many rows are in data frame?
nrow(divvy_bike_trips)  
#Dimensions of the data frame?
dim(divvy_bike_trips)  
#See the first 6 rows of data frame
head(divvy_bike_trips)  
#See list of columns and data types
str(divvy_bike_trips)  
#Statistical summary of data. Mainly for numerics
summary(divvy_bike_trips)  

```

We added more exploitable data for our analysis
```{r Adding usable data}
#Add more exploitable data
#The default format is yyyy-mm-dd
divvy_bike_trips$date <- as.Date(divvy_bike_trips$started_at) 
divvy_bike_trips$month <- format(as.Date(divvy_bike_trips$date), "%m")
divvy_bike_trips$day <- format(as.Date(divvy_bike_trips$date), "%d")
divvy_bike_trips$year <- format(as.Date(divvy_bike_trips$date), "%Y")
divvy_bike_trips$day_of_week <- format(as.Date(divvy_bike_trips$date), "%A")
#The dates are in french
head(divvy_bike_trips)
```

```{r Adding trip length}
#Calculating trip length
divvy_bike_trips$ride_length <- difftime(divvy_bike_trips$ended_at,divvy_bike_trips$started_at)

head(divvy_bike_trips)
```

```{r inspecting new data frame}
#Inspecting our new data frame
str(divvy_bike_trips)
```

```{r converting data types for analysis}
#Converting data types for analysis
is.factor(divvy_bike_trips$ride_length)
 divvy_bike_trips$ride_length <- as.numeric(as.character(divvy_bike_trips$ride_length))
is.numeric(divvy_bike_trips$ride_length)
```

Removing unusable data
```{r removing trip which were not made by customers}
#Removing all trips that were not made by customers
divvy_bike_trips_v2 <- divvy_bike_trips[!(divvy_bike_trips$start_station_name == "HQ QR" | divvy_bike_trips$ride_length<0),]
head(divvy_bike_trips_v2)
```

```{r removing NA values}
#Removing all trips containing NA values
divvy_bike_trips_v3 <- drop_na(divvy_bike_trips_v2)
head(divvy_bike_trips_v3)
```

```{r reviewing ride length}
#Review of the column ride length
summary(divvy_bike_trips_v3$ride_length)
```

## V.Analyzing our dataset
```{r comparing user types}
# Comparing members and casual users
#Mean ride length
aggregate(divvy_bike_trips_v3$ride_length ~ divvy_bike_trips_v3$member_casual, FUN = mean)
#Median ride length
aggregate(divvy_bike_trips_v3$ride_length ~ divvy_bike_trips_v3$member_casual, FUN = median)
#Maximum ride length
aggregate(divvy_bike_trips_v3$ride_length ~ divvy_bike_trips_v3$member_casual, FUN = max)
#Minimum ride length
aggregate(divvy_bike_trips_v3$ride_length ~ divvy_bike_trips_v3$member_casual, FUN = min)
```

```{r calculating average ride by day}
# Calculating the average ride time by each day for members vs casual users
aggregate(divvy_bike_trips_v3$ride_length ~ divvy_bike_trips_v3$member_casual + divvy_bike_trips_v3$day_of_week, FUN = mean)
```

```{r analyzing rider data by type and weekday}
# analyzing ridership data by type and weekday
#creates weekday field using wday()
divvy_bike_trips_v3 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
#groups by usertype and weekday
  group_by(member_casual, weekday) %>%
#calculates the number of rides and average duration   
  summarise(number_of_rides = n()	
# calculates the average duration
            ,average_duration = mean(ride_length)) %>% 	
# sorts
  arrange(member_casual, weekday)								
```

## VI.Visualizations
```{r Visualizing the number of rides bey ride type by day of week}
# Let's visualize the number of rides by rider type
divvy_bike_trips_v3 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title="Divvy bike ride daily rides", subtitle="Daily usage per customer types per day of the week" )
```

```{r Visualizing the average ride length}
# Let's create a visualization for average duration
divvy_bike_trips_v3 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title="Average bike ride duration", subtitle="Per user type and week day")
```

We created a summary of our findings and saved it as a new data frame
```{r Summary of findings}
#Summary of findings
counts <- aggregate(divvy_bike_trips_v3$ride_length ~ divvy_bike_trips_v3$member_casual + divvy_bike_trips_v3$day_of_week, FUN = mean)

#Visualizing our findings
View(counts)
```

### Export findings
```{r Export findings}
#Exporting our findings as a csv file
write.csv(counts, file = 'divvy_bike_trips_length.csv')
```


## VII.Conclusions

Our analysis of our historical data over a period of a year ranging from the month of november 2024 till october 2025, has shed the light on a few interesting facts about our customers us of our services.

Overall for our 2 customer types:

1.Casual riders

*Have an average of rides throughout the week, and we see a gradual rise through out the weekend to reach its peak at an average of on saturdays.

*We also notice casual rides have up to twice the average trip duration throughout the week as compared to member riders

2.Member riders

*Member riders have an average daily usage of bikes during the week which rises up to 2x the amount of trips realized by casual riders from monday to friday

 